Investigating waves and temperature as drivers of kelp morphology

### 2022-08-31 thesis_chapter_2.Rmd
uploaded updated analysis files

---
title: Using numerical modelling to investigate temperature and wave exposure as drivers
  of morphological characteristics of the kelps, Ecklonia maxima and Laminaria pallida
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
output:
  pdf_document: 
    fig_caption: yes
    latex_engine: xelatex
  html_document:
    df_print: paged
  word_document:
    reference_docx: markdown/frontiers_template.docx
bibliography: references.bib
papersize: A4
tables: yes
abstract: Macroalgal morphological variation is determined to a large extent by a combination of environmental factors, with wave exposure and temperature perhaps the main influences, as they are key environmental properties to which a species becomes locally adapted. Macroalgae have shown to exhibit different responses to different magnitudes of exposure to waves, such as reduction in overall size and strength increasing traits. In terms of temperature, warmer environments have been shown to reduce the overall size of resident and transplanted species. However, none of the past studies have identified specific wave and temperature metrics responsible for the morphological adaptation macroalgae exhibit. Past research has often used simple or two-dimensional models of wave exposure, which do not take into account important aspects of the nearshore environment such as wave breaking, refraction and diffraction. Furthermore, past studies have often used satellite-derived datasets as sources for temperature data, however, such data have been shown to have large bias when applied to the nearshore environment. This study used in situ temperature data and wave power metrics calculated from a 3D-numerical model to identify specific temperature and wave metrics responsible for morphological adaptation of the kelp, Ecklonia maxima and Laminaria pallida. Between temperature and wave exposure, the results identify wave exposure as the main influencer of morphological adaptation while identifying specific wave metrics. Furthermore, the results show differences in wave metrics between species; and between deep and shallow populations.'

@@

```{r include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(fig.align = "center")
```
# Introduction

Kelps are a group of large seaweeds of the order Laminariales (Ochrophyta), which despite their relatively low taxonomic diversity of species in genera [@Bolton2010], form the basis of one of the most productive ecosystems globally [@Mann1973; @Krumhansl2012]. Kelps generally have a dependence on cool-temperate and arctic seawater temperatures [@Bolton2010; @Mohring2014-bk; @Cavanaugh2011-ib; @Dayton1985], and dominate the nearshore biomass within the rocky shallow coasts in both hemispheres [@Steneck2002]. Their size and complex morphology provide a heterogeneous habitat structure [@Steneck2002] that accommodate a multitude of turf and sub-canopy seaweed species, and diverse assemblages of sessile and mobile invertebrates and vertebrates [@Mann1973; @Steneck2002], each depending on a wide suite of ecological services provided by kelp forests [@Dayton1985; @Gaines1987-bg; @Paul1993-bm; @Levin1994-qf; @Anderson1997; @Willis2003-xo]. Wave exposure and temperature are regarded as important environmental drivers of kelp forests, and play a role in their distribution [@Gorman2013-nj], abundance [@Dayton1998; @Cavanaugh2011-ib], diversity [@Wing2007-tv; @Wernberg2008], composition [@Dayton1985; @Leliaert2000a; @Harley2012; @Norderhaug2012-ks], growth [@Cousens1982-ck] and productivity [@Dayton1998;@Krumhansl2012; @Pedersen2012].

Given that kelps are cool-temperate organisms which are vulnerable to dislodgement under high wave exposure scenarios, increases in storm frequency and magnitude, and changes in ocean temperature, pose a direct threat to their survival; which is a concern considering that these variables are known to be affected by climate change [@Russo2014; @Reguero2019; @Weisse2010; @Rose2012; @Rummukainen2012]. Warmer temperatures reduces the resilience of kelp individuals to storm disturbance [@Wernberg2010b], causes fragmentation through weakening tissues [@Simonson2015] and reduces growth and productivity [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]; while storms and high wave energy dislodge and break kelps [@byrnes2011; @graham2004; @seymour1989]. Despite these threats and disturbances kelps persist across a broad range of environments which is largely due to their morphological plasticity [@FowlerWalker2006; @Wernberg2010a]. A study by [@Wernberg2003-mp] investigated the morphology of Ecklonia radiata (C.Agardh) J.Agardh in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australian coast. They found no correlation between spatial distance and morphological similarity; rather the morphology of the kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales [@Wernberg2003-mp].

Although multiple environmental factors play a role in kelp morphological adaptation, wave exposure and temperature have been identified as the main influencers across various kelp species [@Hurd2000;@Denny2002;@Thomsen2004;@FowlerWalker2005;@Wernberg2005c;@mabin2013;@Bekkby2014]. For example, a study by  @Serisawa2002-d compared the morphology of Ecklonia cava Kjellman growing at warmer and cooler sites, and showed wrinkling of the blade and reduced size to be a characteristic of higher temperature ranges. Warmer temperature affects important kelp physiological processes such as photosynthesis and respiration, which influences growth and productivity and ultimately leads to a reduction in size [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]. Past research has shown that in highly wave exposed areas kelp morphology tends to take on characteristics which reduce drag, increase strength and increase flexibility [@Denny2002; @Hurd2000; @FowlerWalker2005; @Thomsen2004]. For example, a study by @Wernberg2005c examined the consistency of wave exposure as a driver of E. radiata across a broad geographic range, and showed trends toward drag-reducing (small size, narrow laterals and blades, low spinosity) and increased strength (large holdfast, thick stipe and thick blades and lamina) at high wave exposure sites.

Measures of wave exposure in ecological studies often incorporate integrative measures of hydrodynamic conditions at a particular site which are based on cartographical models of exposure. Cartographical models of wave exposure are “fetch-based-models” which measure the length of open water associated with a particular site in a straight line, and are regarded as simple measures of wave exposure. Advances in numerical modeling based on physical/linear wave theory incorporate more complexity (wind forcing, wave-wave interactions, wave breaking, diffraction and variation in wave direction) into the models and allow for a quantitative, reproducible approach for characterizing the hydrodynamic environment. Although @mabin2013 investigated kelp morphological characteristics using a 2D WAM model, currently no studies investigating macroalgal morphological characteristics in relation to the wave environment have incorporated 3D spectral numerical modeling. Furthermore no studies have identified specific wave metrics that drive kelp morphological characteristics, which the current work will aim to achieve. Other considerations such as possible differences in morphology between sub-canopy and canopy species, and shallow and deep water kelp populations have not been considered using advanced numerical and statistical tools. Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Deep and shallow water environments may differ in abiotic processes such as the degree of water column mixing [@Smit2013a], solar heating [@Dellatorre2012; @Dunne2001], and effects of wave dampening [@Kobayashi1993-os; @Mia2013; @Dubi1995].

The kelps *Laminaria pallida* Greville and *Ecklonia maxima* (Osbeck) Papenfuss are important habitat forming seaweeds that exist around the coast of South Africa and offer a unique opportunity to investigate the drivers of macroalgal morphological characteristics between canopy and sub-canopy species, as well as between deep and shallow water environments. Although both these species exist in the sub-tidal, *L. pallida* dominates deeper waters while *E. maxima* forms dense surface canopies in the shallow sub-littoral. Therefore, E. maxima may be exposed to variations in wind driven surface waves, swell waves and temperature, while *L. pallida* is exposed to variations in swell waves, wave exposure and temperature due to differences in depth. @Molloy1996-la investigated the effect of depth and wave exposure on *L. pallida* and showed that depth had a greater effect than wave exposure when considering all the morphological characteristics; when considering individual characteristics, however, wave exposure had the most significant effect on blade thickness. Another study by @Rothman2017-pn investigated the changes in morphology in shallow populations of *L. pallida* and *E. maxima* along the South African coastline and into Namibia; *E. maxima* exhibited no morphological changes along the coast but the stipes of *L. pallida* become increasingly hollow further north along the coastline. The authors suggested that turbidity in relation to light attenuation was the environmental driver responsible for this change.

The aim of this study is, therefore, to understand how temperature and wave exposure can influence the morphology of sub-canopy and canopy kelps, as well as between deep and shallow populations of the same species. The current study will also aim to identify specific temperature and wave metrics which drive kelp morphological characteristics. This will be achieved by initially understanding the variation in temperature and waves using a numerical model. The numerical model will then be used as a basis for investigating the consequences for morphological characteristics of *E. maxima* and *L. pallida* using advanced statistical tools.

@@

Sites were chosen to represent an array of temperature and wave gradients, as indicated in Figure \ref{fig:map}. St. Helena Bay and Betty’s Bay constituted the north-western and south-eastern boundary sites, respectively. These sites are roughly 300 km apart, lie within separate marine provinces  @Smit2017, and span the majority of the south-west coast, in varying thermal and wave energy regimes. The region is dominated by kelp communities that persist in contrasting abiotic environments. The west coast region has been termed cool-temperate, which is defined as a region where mean monthly temperatures are always above 10°C and below 15°C [@Smit2013a]. East of Cape Point marks the beginning of an overlap or transition area, which is also referred to as the Benguela-Agulhas Transition Zone [@Smit2013a]. The Agulhas Marine Province is characterized by a wide temperature range of up to 7°C difference between mean monthly temperatures between summer and winter and is classified as warm-temperate [@Smit2013a].

```{r map processing, message=FALSE, warning=FALSE, include=FALSE}
## Load software:
library(ggplot2)
library(rgdal)
library(tidyverse)
library(viridis)
library(gridExtra)
source("map/scaleBarFunc.R")
## Coastline of Southern Africa
load("map/south_africa_coast_hi_res.Rdata")
## Read in metadata
study_sites <- read.csv("map/metadata_map.csv", sep = ";")

@@

The annual and seasonal patterns in both wind and swell direction influence the thermal regime around the coastline by inducing upwelling. Upwelling brings deep, cool nutrient-rich water to coastal areas which causes a decrease in temperature [@Field1980a; @Andrews1980; @Gill1974-nt; @Cram1970-jw; @Rouault2010-gi; @Blanke2002]. The wind speed and wind direction help drive upwelling in summer on the western side of the Cape Peninsula, where southerly winds blow parallel to the coast and trigger upwelling events [@Field1980a; @Rouault2010-gi]. This is not the case for False Bay, which is shielded from the dominating wind and swell that the Cape Peninsula is subjected to. In the summer months, upwelling becomes more favorable in False Bay due to the change in swell and wind direction, causing lower temperatures to dominate during this time [@Rouault2010-gi; @Andrews1980]. Due to the Cape Peninsula temperate latitude, winter months bring an increased frequency of frontal depressions that originate from the Southern Ocean [@Reason2006]. These low pressures are joined by large swells with increased wave energy. The nearshore environment, with the accompanied biota, therefore experiences high wave energy events, with increased frequency in winter [@Veitch2019]. The large peninsula acts as an obstruction for large south westerly swells, providing decreased wave energy along the west side of False Bay [@Shipley1964]. Conversely, the west coast of Cape Point is battered by these large swells. Multiple sites, therefore, exist where kelps grow in diverse temperature and wave energy climates, in close proximity and on the same side. The topography and elevation along the Cape Peninsula channel and shield winds along False Bay. This is, however, absent in winter, when strong northerly winds are prevalent from St. Helena Bay to Betty’s Bay [@Field1980a; @Jury1985-nr; @Andrews1980; @Jury1980-ov].

@@

Shallow water temperatures were sourced from The South African Coastal Temperature Network (SACTN) website1. In terms of nearshore temperature, in situ data are preferred over satellite sea surface temperature (SST), which have shown to exhibit large biases [@Smit2013a]. The underwater temperature recorders (UTRs) are attached to concrete mooring blocks or railway line sections deployed in the shallow at ∼3 m depth. The UTRs were comprised of Starmon Mini recorders (Star-Oddi, Reykjavik, Iceland/Onset Hobo® U22-001) accurate to ±0.2°C. All in situ temperature data were reduced to a time-coded, continuous series of monthly means over various time-scales [@Smit2013a]. It was decided that the annual time-scale would be used for analysis. Linear interpolated SST were calculated for sites where in situ recorders were absent. These data were used to group sites into “Cool temperate” and “Warm temperate” categorizations to investigate the possible effect of thermal regime on kelp morphological characteristics.

@@

Wave model variable data formed part of the South African Coastal Vulnerability Assessment, presented to the Department of Environmental Affairs (DEA) and produced by the Council for Scientific and Industrial Research (CSIR) [@Rautenbach2015a]. The South African wave climate was modeled via 20 spectral numerical wave models that simulated the offshore wave climate to the nearshore. The boundary conditions of these models were obtained by using the NOAA Wave Watch III (WWIII) model output, distributed via the National Centers for Environmental Prediction (NCEP) product (Office of the Director 2000; Environmental Modeling Center/Marine Modeling Branch 2005). The particular hindcast product utilized during the DEA-CSIR study spans 1994–2013 at a 3-hour resolution. These data were then used to model swell propagation into the coastal models while wind waves (seas) were generated via stationary computations in the Simulating Waves in the Nearshore SWAN model [@Booij1997-nx]. The assumption of stationary computations is acceptable as the model domains were small enough so the temporal variation of the model boundary was slower than the time it takes for that boundary condition to propagate to the coast. SWAN allows one to extract wave variables from specific gridded locations in the nearshore. For False Bay, a resolution of 200 meters was modeled and output produced at both the 7 m and 15 m isobaths. A 200 m resolution was used as the False Bay computational grid was nested within a larger 1 km resolution grid. This allowed for a computational effective wave resolution of increasing resolution, from the NCEP, low resolution output to nested, high resolution coastal output. For Table Bay and east of Cape Hangklip the resolution was 500 m and also had output at the 7 and 15 m isobaths. These contour outputs were chosen in the original study by the CSIR as most engineering run-up calculations require wave parameter information at these contour depths and were the main focus of the original study. For this study the 7 m contours were used. These data were then used to calculate over all wave power (kW/m) as this was considered the best measure of overall wave exposure. Annual wave power was plotted and categorized into different wave exposure categories which ranged from fully sheltered to extremely exposed.

@@

The morphological characteristics of both species are presented in figure \ref{fig:schematic}. *Laminaria pallida* is characterized by a single smooth blade which is divided longitudinally into sections, and develops from a single meristematic region located at the junction between the blade and the stipe [@Dyer2018-rv]. This species has a solid stipe along the south-west coast but develops a hollow stipe along the west coast northward and into Namibia [@Rothman2017-pn]. *Ecklonia maxima* consists of a single primary blade which develops above a gas-filled float and a hollow stipe below. Secondary blades are produced laterally from the primary blade from several meristematic regions along the margins of the primary blade [@Dyer2018-rv]. Both species are held to the substrate by finger-like haptera, collectively known as the holdfast.

@@

Morphological measurements of L. pallida and E. maxima were collected at 18 sites along the Western Cape coast of South Africa (Figure \ref{fig:map}) between October 2014 and April 2015. AlgaeBase was used to confirm the latest taxonomic associations for each species [@Guiry2019]. Divers sampled kelp individuals by means of SCUBA for deeper populations and freediving for shallower populations. The sampled individuals were then taken ashore and the relevant morphological measurements taken. In each instance only the 13 largest individuals were sampled to ensure that only mature and fully grown sporophytes were measured. The age of *E. maxima* was determined by size, presence at the surface as well as a darker coloration of the stipe; as this occurs with age. In the case of *L. pallida*, the individuals with the longest stipe were selected which is considered a reasonable estimate of age. In each case the age of the individuals was estimated visually before the individual was sampled. In instances where one of the species was not present a site, the nearest population of the missing species was sampled. Because the macroalgae differ in morphological features, species-specific morphological characteristics were included and are presented in Table 1. The stipe length for *E. maxima* was considered to be the portion of the plant just below the pneumatocyst and just above the holdfast. Stipe circumference was measured just above the holdfast for both species. Between February 2017 and November 2018, morphological measurements for E. maxima individuals in shallow water (<1 m) at seven sites along the Western Cape coast of South Africa were also collected. The same morphological characteristic measurements were taken as for the deeper *E. maxima*, and this allowed comparison between morphological characteristics between deep and shallow individuals within sites. Measurements were not collected for *L. pallida* in the shallow depths, as this species is largely absent from the shallow in this portion of the South African coastline.

@@

```
In general, the mean temperature increases for sites located within False Bay with less variation around the mean temperature compared to sites on the western side of the peninsula (see Figure \ref{fig:Temperature variables plot}A). This is verified by statistical difference based on region, which shows a significantly lower (p < 0.05, Figure \ref{fig:Temperature variables plot}B) mean temperature and less variation in temperature for sites located on the western side of the peninsula compared to sites on the eastern side.

@@

Six out of the ten deep *E. maxima* morphological characteristics showed differences between cool temperate and warm temperate regime except for frond mass, frond length, stipe mass and epiphyte length. Mean values for primary length, primary width and stipe circumference were significantly higher for kelp from the cool temperate regime than for kelp from the warm temperate regime (see Figure \ref{fig:{deep eck temp plot}}). The remaining morphological characteristics, stipe length, number of tufts and total length had significantly lower means for kelp from cool temperate compared to kelp from the warm temperate regime (p < 0.05; see Figure 8). Significant differences amongst wave exposure categories were found for all morphological characteristics of deep E. maxima populations (see Figure \ref{fig:deep eck temp plot}
). Certain morphological characteristics showed a gradual increase in mean value with increasing degree of wave exposure. The remaining morphological characteristics exhibited decreased means for only the higher wave exposure categories, followed by a sharp increase in mean value for the highest wave exposure category.

>>>>

---
title: 'Investigating waves and temperature as drivers of kelp morphology'
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
output:
  pdf_document: 
    fig_caption: yes
    latex_engine: tectonic
  html_document:
    df_print: paged
  word_document:
    reference_docx: markdown/thesis_template.docx
bibliography: references.bib
papersize: A4
tables: yes
abstract: 'Macroalgal morphological variation is determined to a large extent by a combination of environmental factors, with wave exposure and temperature perhaps the main influences, as they are key environmental properties to which a species becomes locally adapted. Macroalgae have shown to exhibit different responses to different magnitudes of exposure to waves, such as reduction in overall size and strength increasing traits. In terms of temperature, warmer environments have been shown to reduce the overall size of resident and transplanted species. However, none of the past studies have identified specific wave and temperature metrics responsible for the morphological adaptation macroalgae exhibit. Past research has often used simple or two-dimensional models of wave exposure, which do not take into account important aspects of the nearshore environment such as wave breaking, refraction and diffraction. Furthermore, past studies have often used satellite-derived datasets as sources for temperature data, however, such data have been shown to have large bias when applied to the nearshore environment. This study used in situ temperature data and wave power metrics calculated from a 3D-numerical model to identify specific temperature and wave metrics responsible for morphological adaptation of the kelp, Ecklonia maxima and Laminaria pallida. Between temperature and wave exposure, the results identify wave exposure as the main influencer of morphological adaptation while identifying specific wave metrics. Furthermore, the results show differences in wave metrics between species; and between deep and shallow populations.'

@@

```{r include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(fig.align = "center")
```

# Introduction

Kelps are a group of large seaweeds of the order Laminariales (Ochrophyta), which despite their relatively low taxonomic diversity of species in genera [@Bolton2010], form the basis of one of the most productive ecosystems globally [@Mann1973; @Krumhansl2012]. Kelps generally have a dependence on cool-temperate and arctic seawater temperatures [@Bolton2010; @Mohring2014-bk; @Cavanaugh2011-ib; @Dayton1985], and dominate the nearshore biomass within the rocky shallow coasts in both hemispheres [@Steneck2002]. Their size and complex morphology provide a heterogeneous habitat structure [@Steneck2002] that accommodate a multitude of turf and sub-canopy seaweed species, and diverse assemblages of sessile and mobile invertebrates and vertebrates [@Mann1973; @Steneck2002], each depending on a wide suite of ecological services provided by kelp forests [@Dayton1985; @Gaines1987-bg; @Paul1993-bm; @Levin1994-qf; @anderson1997; @Willis2003-xo]. Wave exposure and temperature are regarded as important environmental drivers of kelp forests, and play a role in their distribution [@Gorman2013-nj], abundance [@Dayton1998-xw; @Cavanaugh2011-ib], diversity [@Wing2007-tv; @Wernberg2008], composition [@Dayton1985; @Leliaert2000a; @Harley2012; @Norderhaug2012-ks], growth [@Cousens1982-ck] and productivity [@Dayton1998-xw;@Krumhansl2012; @Pedersen2012a].

Given that kelps are cool-temperate organisms which are vulnerable to dislodgement under high wave exposure scenarios, increases in storm frequency and magnitude, and changes in ocean temperature, pose a direct threat to their survival; which is a concern considering that these variables are known to be affected by climate change [@Weisse2010;@Rose2012; @Rummukainen2012;@Russo2014; @Reguero2019]. Warmer temperatures reduces the resilience of kelp individuals to storm disturbance [@Wernberg2010b], causes fragmentation through weakening tissues [@Simonson2015] and reduces growth and productivity [@zimmerman1986; @gerard1997; @bearham2013; @gao2013]; while storms and high wave energy dislodge and break kelps[@seymour1989; @graham2004; @byrnes2011]. Despite these threats and disturbances kelps persist across a broad range of environments which is largely due to their morphological plasticity [@FowlerWalker2006; @Wernberg2010a]. A study by @Wernberg2003-mp investigated the morphology of *Ecklonia radiata* (C.Agardh) J.Agardh in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australian coast. They found no correlation between spatial distance and morphological similarity; rather the morphology of the kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales [@Wernberg2003-mp].

Although multiple environmental factors play a role in kelp morphological adaptation, wave exposure and temperature have been identified as the main influencers across various kelp species [Hurd2000; @Denny2002a; @Thomsen2004; @FowlerWalker2005; @Wernberg2005c; @mabin2013; @bekkby2014]. For example, a study by @Serisawa2002-dc compared the morphology of *Ecklonia cava* Kjellman growing at warmer and cooler sites, and showed wrinkling of the blade and reduced size to be a characteristic of higher temperature ranges. Warmer temperature affects important kelp physiological processes such as photosynthesis and respiration, which influences growth and productivity and ultimately leads to a reduction in size [@zimmerman1986; @gerard1997; @bearham2013; @gao2013]. Past research has shown that in highly wave exposed areas kelp morphology tends to take on characteristics which reduce drag, increase strength and increase flexibility (@Hurd2000; @Denny2002a; @Thomsen2004; @FowlerWalker2005]. For example, a study by @Wernberg2005c examined the consistency of wave exposure as a driver of *E. radiata* across a broad geographic range, and showed trends toward drag- reducing (small size, narrow laterals and blades, low spinosity) and increased strength (large holdfast, thick stipe and thick blades and lamina) at high wave exposure sites.                                
Measures of wave exposure in ecological studies often incorporate integrative measures of hydrodynamic conditions at a particular site which are based on cartographical models of exposure. Cartographical models of wave exposure are “fetch-based-models” which measure the length of open water associated with a particular site in a straight line, and are regarded as simple measures of wave exposure. Advances in numerical modeling based on physical/linear wave theory incorporate more complexity (wind forcing, wave-wave interactions, wave breaking, diffraction and variation in wave direction) into the models and allow for a quantitative, reproducible approach for characterizing the hydrodynamic environment. Although @mabin2013 investigated kelp morphological characteristics using a 2D WAM model, currently no studies investigating macroalgal morphological characteristics in relation to the wave environment have incorporated 3D spectral numerical modeling. Furthermore no studies have identified specific wave metrics that drive kelp morphological characteristics, which the current work will aim to achieve. Other considerations such as possible differences in morphology between sub-canopy and canopy species, and shallow and deep water kelp populations have not been considered using advanced numerical and statistical tools. Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Deep and shallow water environments may differ in abiotic processes such as the degree of water column mixing [@Smit2013a], solar heating [@Dunne2001; @Dellatorre2012], and effects of wave dampening [@Kobayashi1993-os; @Dubi1995; @ma2013].                        
The kelps *Laminaria pallida* Greville and *Ecklonia maxima* (Osbeck) Papenfuss are important habitat forming seaweeds that exist around the coast of South Africa and offer a unique opportunity to investigate the drivers of macroalgal morphological characteristics between canopy and sub-canopy species, as well as between deep and shallow water environments. Although both these species exist in the sub-tidal, *L. pallida* dominates deeper waters while *E. maxima* forms dense surface canopies in the shallow sub-littoral. Therefore, *E. maxima* may be exposed to variations in wind driven surface waves, swell waves and temperature, while *L. pallida* is exposed to variations in swell waves, wave exposure and temperature due to differences in depth. @Molloy1996-la investigated the effect of depth and wave exposure on *L. pallida* and showed that depth had a greater effect than wave exposure when considering all the morphological characteristics; when considering individual characteristics, however, wave exposure had the most significant effect on blade thickness. Another study by @Rothman2017-pn investigated the changes in morphology in shallow populations of *L. pallida* and *E. maxima* along the South African coastline and into Namibia; _E. maxima _exhibited no morphological changes along the coast but the stipes of  *L.pallida* become increasingly hollow further north along the coastline. The authors suggested that turbidity in relation to light attenuation was the environmental driver responsible for this change.

The aim of this study is, therefore, to understand how temperature and wave exposure can influence the morphology of sub-canopy and canopy kelps, as well as between deep and shallow populations of the same species. The current study will also aim to identify specific temperature and wave metrics which drive kelp morphological characteristics. This will be achieved by initially understanding the variation in temperature and waves using a numerical model. The numerical model will then be used as a basis for investigating the consequences for morphological characteristics of *E. maxima* and *L.pallida* using advanced statistical tools.

@@

Sites were chosen to represent an array of temperature and wave gradients, as indicated in Figure 1. St. Helena Bay and Betty’s Bay constituted the north-western and south-eastern boundary sites, respectively. These sites are roughly 300 km apart, lie within separate marine provinces [@Smit2017], and span the majority of the south-west coast, in varying thermal and wave energy regimes. The region is dominated by kelp communities that persist in contrasting abiotic environments. The west coast region has been termed cool-temperate, which is defined as a region where mean monthly temperatures are always above 10◦C and below 15◦C  [@Smit2013a. East of Cape Point marks the beginning of an overlap or transition area, which is also referred to as the Benguela-Agulhas Transition Zone [@Smit2013a. The Agulhas Marine Province is characterized by a wide temperature range of up to 7◦C difference between mean monthly temperatures between summer and winter and is classified as warm-temperate  [@Smit2013a.

```{r map processing, message=FALSE, warning=FALSE, include=FALSE}
## Load software:
library(ggplot2)
library(tidyverse)
library(viridis)
library(gridExtra)
source("map/scaleBarFunc.R")
## Coastline of Southern Africa
load("map/south_africa_coast_hi_res.Rdata")
## Read in metadata
study_sites <- read.csv("map/metadata_map.csv", sep = ";")

@@

The annual and seasonal patterns in both wind and swell direction influence the thermal regime around the coastline by inducing upwelling. Upwelling brings deep, cool nutrient-rich water to coastal areas which causes a decrease in temperature [@Field1980a; @Andrews1980; @Gill1974-nt; @Cram1970-jw; @Rouault2010-gi; @blanke2002]. The wind speed and wind direction help drive upwelling in summer on the western side of the Cape Peninsula, where southerly winds blow parallel to the coast and trigger upwelling events [@Field1980a; @Rouault2010-gi]. This is not the case for False Bay, which is shielded from the dominating wind and swell that the Cape Peninsula is subjected to. In the summer months, upwelling becomes more favorable in False Bay due to the change in swell and wind direction, causing lower temperatures to dominate during this time [@Rouault2010-gi; @Andrews1980]. Due to the Cape Peninsula temperate latitude, winter months bring an increased frequency of frontal depressions that originate from the Southern Ocean [@Reason2006]. These low pressures are joined by large swells with increased wave energy. The nearshore environment, with the accompanied biota, therefore experiences high wave energy events, with increased frequency in winter [@Veitch2019]. The large peninsula acts as an obstruction for large south westerly swells, providing decreased wave energy along the west side of False Bay [@Shipley1964]. Conversely, the west coast of Cape Point is battered by these large swells. Multiple sites, therefore, exist where kelps grow in diverse temperature and wave energy climates, in close proximity and on the same side. The topography and elevation along the Cape Peninsula channel and shield winds along False Bay. This is, however, absent in winter, when strong northerly winds are prevalent from St. Helena Bay to Betty’s Bay [@Field1980a; @Jury1985-nr; @Andrews1980; @Jury1980-ov].

@@

Shallow water temperatures were sourced from The South African Coastal Temperature Network (SACTN) website1. In terms of nearshore temperature, *in situ* data are preferred over satellite sea surface temperature (SST), which have shown to exhibit large biases [@Smit2013a]. The underwater temperature recorders (UTRs) are attached to concrete mooring blocks or railway line sections deployed in the shallow at ∼3 m depth. The UTRs were comprised of *Starmon Mini* recorders (_Star-Oddi_, Reykjavik, Iceland/_Onset Hobo_⃝R U22-001) accurate to ±0.2◦C. All *in situ* temperature data were reduced to a time-coded, continuous series of monthly means over various time-scales [@Smit2013a]. It was decided that the annual time-scale would be used for analysis. Linear interpolated SST were calculated for sites where *in situ* recorders were absent. These data were used to group sites into “Cool temperate” and “Warm temperate” categorizations to investigate the possible effect of thermal regime on kelp morphological characteristics.

@@

Wave model variable data formed part of the South African Coastal Vulnerability Assessment, presented to the Department of Environmental Affairs (DEA) and produced by the Council for Scientific and Industrial Research (CSIR) [@Rautenbach2015a]. The South African wave climate was modeled _via _20 spectral numerical wave models that simulated the offshore wave climate to the nearshore. The boundary conditions of these models were obtained by using the NOAA Wave Watch III (WWIII) model output, distributed via the National Centers for Environmental Prediction (NCEP) product (Office of the Director 2000; Environmental Modeling Center/Marine Modeling Branch 2005). The particular hindcast product utilized during the DEA-CSIR study spans 1994–2013 at a 3-hour resolution. These data were then used to model swell propagation into the coastal models while wind waves (seas) were generated via stationary computations in the Simulating Waves in the Nearshore SWAN model [@Booij1997-nx]. The assumption of stationary computations is acceptable as the model domains were small enough so the temporal variation of the model boundary was slower than the time it takes for that boundary condition to propagate to the coast. SWAN allows one to extract wave variables from specific gridded locations in the nearshore. For False Bay, a resolution of 200 meters was modeled and output produced at both the 7 m and 15 m isobaths. A 200 m resolution was used as the False Bay computational grid was nested within a larger 1 km resolution grid. This allowed for a computational effective wave resolution of increasing resolution, from the NCEP, low resolution output to nested, high resolution coastal output.

For Table Bay and east of Cape Hangklip the resolution was 500 m and also had output at the 7 and 15 m isobaths. These contour outputs were chosen in the original study by the CSIR as most engineering run-up calculations require wave parameter information at these contour depths and were the main focus of the original study. For this study the 7 m contours were used. These data were then used to calculate over all wave power (kW/m) as this was considered the best measure of overall wave exposure. Annual wave power was plotted and categorized into different wave exposure categories which ranged from fully sheltered to extremely exposed.

@@

The morphological characteristics of both species are presented in Figure 2. *Laminaria pallida* is characterized by a single smooth blade which is divided longitudinally into sections, and develops from a single meristematic region located at the junction between the blade and the stipe [@Dyer2018-rv]. This species has a solid stipe along the south-west coast but develops a hollow stipe along the west coast northward and into Namibia [@Rothman2017-pn]. *Ecklonia maxima* consists of a single primary blade which develops above a gas-filled float and a hollow stipe below. Secondary blades are produced laterally from the primary blade from several meristematic regions along the margins of the primary blade [@Dyer2018-rv]. Both species are held to the substrate by finger-like haptera, collectively known as the holdfast.

@@

Morphological measurements of *L.pallida* and _E. maxima_ were collected at 18 sites along the Western Cape coast of South Africa (Figure 1) between October 2014 and April 2015. AlgaeBase was used to confirm the latest taxonomic associations for each species (Guiry and Guiry, 2019). Divers sampled kelp individuals by means of SCUBA for deeper populations and freediving for shallower populations. The sampled individuals were then taken ashore and the relevant morphological measurements taken. In each instance only the 13 largest individuals were sampled to ensure that only mature and fully grown sporophytes were measured. The age of *E. maxima* was determined by size, presence at the surface as well as a darker coloration of the stipe; as this occurs with age. In the case of *L.pallida*, the individuals with the longest stipe were selected which is considered a reasonable estimate of age. In each case the age of the individuals was estimated visually before the individual was sampled. In instances where one of the species was not present at a site, the nearest population of the missing species was sampled. Because the macroalgae differ in morphological features, species-specific morphological characteristics were included and are presented in Table 1. The stipe length for *E. maxima* was considered to be the portion of the plant just below the pneumatocyst and just above the holdfast. Stipe circumference was measured just above the holdfast for both species. Between February 2017 and November 2018, morphological measurements for *E. maxima* individuals in shallow water (<1m) at seven sites along the Western Cape coast of South Africa were also collected. The same morphological characteristic measurements were taken as for the deeper *E. maxima*, and this allowed comparison between morphological characteristics between deep and shallow individuals within sites. Measurements were not collected for *L.pallida* in the shallow depths, as this species is largely absent from the shallow in this portion of the South African coastline.

```{r table 2.1, message=FALSE, warning=FALSE, echo=FALSE}

library(flextable)

morph_table <- read.csv("tables/chapter_2/table_1.csv", sep = ";")
morph_table <- flextable(data = morph_table)
morph_table <- theme_zebra(morph_table, odd_header = "transparent", even_header = "transparent")
morph_table <- set_header_labels(morph_table, Morphological.Characteristic = "Morphological Characteristic", Unit.measurement = "Unit.measurement")
morph_table <- align(x= morph_table, align = "center", part = "all")
morph_table <- set_table_properties(morph_table, width = 1, layout = "autofit")
morph_table <- set_caption(morph_table, "A list of measured environmental variables (temperature, waves and wind) and their associated abbreviations.")
morph_table

@@

In general, the mean temperature increases for sites located within False Bay with less variation around the mean temperature compared to sites on the western side of the peninsula (see Figure \ref{fig:Temperature variables plot}A). This is verified by statistical difference based on region, which shows a significantly lower (p < 0.05, Figure \ref{fig:Temperature variables plot}B) mean temperature and less variation in temperature for sites located on the western side of the peninsula compared to sites on the eastern side.

@@

Six out of the ten deep *E. maxima* morphological characteristics showed differences between cool temperate and warm temperate regime except for frond mass, frond length, stipe mass and epiphyte length. Mean values for primary length, primary width and stipe circumference were significantly higher for kelp from the cool temperate regime than for kelp from the warm temperate regime (see Figure \ref{fig:{deep eck temp plot}}). The remaining morphological characteristics, stipe length, number of tufts and total length had significantly lower means for kelp from cool temperate compared to kelp from the warm temperate regime (p < 0.05; see Figure 8). Significant differences amongst wave exposure categories were found for all morphological characteristics of deep *E. maxima* populations (see Figure \ref{fig:deep eck temp plot}
). Certain morphological characteristics showed a gradual increase in mean value with increasing degree of wave exposure. The remaining morphological characteristics exhibited decreased means for only the higher wave exposure categories, followed by a sharp increase in mean value for the highest wave exposure category.
